---
title: "Ecosystem Service Climate Regulation in Cities | Norway"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
---

```{css, echo=FALSE}
table {
  float: left;
}
```

Magnus Merkle (SSB)

Naturregnskapsprosjekt, 18. August 2026

\

# Changelog

Changes since version 0:

- First version

# Abstract and Limitation

In this notebook we quantify the ecosystem service of local climate regulation (climate regulation in cities) for the Norwegian mainland in 2024. Most operations are executed with functions from R packages dplyr and tidyr (for general data formatting), terra (for geospatial data analysis), and stats (for regressions). The notebook is created on the basis of the following sources:

- INCA GIS Plugin Code Repository (2026). Last accessed 13 August 2026, the relevant script is here: <https://code.europa.eu/inca/gis-plugins/gis-plugins/-/blob/main/src/inca/localclimate/__init__.py>
- Marando et al. (2022) Urban heat island mitigation by green infrastructure in European Functional Urban Areas. Sustainable Cities and Society, 103564
- Eurostat (2026) Ecosystem accounts handbook, 2026 edition
- Immerzeel & Korkou (2024) Evaluation of the INCA-Tool for the purpose of ecosystem services accounting in Norway. EU Grants Project 2022-NO-EGD- Work Package 2. Norwegian Institute for Nature Research

The quantification process follows the Eurostat Ecosystem accounts handbook.

# 1. Filepaths and Functions

The main functions we use are part of R packages dplyr and tidyr, terra, and stats.

```{r, results = 'hide', message = FALSE, warning=FALSE}
# make space
rm(list = ls())
# packages
library(dplyr)
library(tidyr)
library(terra)
library(sf)
library(knitr)
library(viridisLite)
library(here)
library(stats)
library(tinytex)
# filepath
dfp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "data","INCA")
# file path to results
rfp<-file.path(here::here(),"results")
# A function to extract levels vectors from factor data stored in rasters 
# rst is raster file, lrn is layer name
gc()
```

# 2. Data Sources

Here is a table of data sources used in this project.

```{r}
dfs<-data.frame(matrix(nrow=10,ncol=8))
colnames(dfs) <- c("Input to","Data","Name","Provider","Resolution","Year","Future Updates","Notes")

dfs[1,]<-c("Estimation and simulation",
           "Reference land cover map (LCM) with lookup table",
           "Grunnkart_OT",
           "Kartverket, NIBIO, SSB, Mdir",
           "100m grid",
           "2024",
           "2025",
           "This is probably the most important input because we derive ecosystems, curve numbers, and economic assets from it. Updates should ideally be used to quantify changes in service flows over time. Comparability over time would be relevant for this purpose.")
dfs[2,]<-c("Estimation and simulation",
           "City shapefiles",
           "....",
           ".....",
           "Vector",
           "...",
           "...",
           "The delineation of city boundaries is important because it defines how much of the surrounding greenspace is taken into account.")
dfs[3,]<-c("Estimation and simulation",
           "Local surface temperature data",
           "...",
           "Eurostat (INCA default version 2026)",
           ".... grid",
           "....",
           "....",
           "....")
dfs[4,]<-c("Estimation",
           "Evaporation ...",
           "....",
           "...",
           "...",
           "...",
           "...",
           "...")
dfs[5,]<-c("Estimation",
           "Weather station data",
           "...",
           "...",
           "not gridded",
           "...",
           "...",
           "...")
#print
kable(dfs)
# clean up
rm(dfs)
```

# 3. Estimate a Model of LST for each City

## 3.1 Overview

In this first step we use gridded data to estimate a linear model of local surface temperature (LST) predicted by tree cover density (TCD) and evaporation. We estimate this model for each city, so it requires that our gridded data is cropped to city boundaries plus a buffer of 150 metres (as specified in Eurostat's handbook). The model is: 
$$ LST_{j} = \alpha_{1,j} + \beta_{1} TCD + \gamma_{1} EVAP $$ 
For each city, we save the estimated intercept of the linear model $\widehat{\alpha_{1,j}}$, which is equal to the LST in a scenario with no trees and evaporation. This is a purely hypothetical number with no interpretative value. It is needed as an intermediate step.

## 3.2 Import data

Read in city shapefiles, LST raster, TCD raster, EVAP raster. We also import Norwegian Land Cover data and some Open Street Map data, for orientation purpose. Then crop all raster files to the Norwegian city boundaries with a 150m buffer. 

We start by identifying cities in Norway that are counted within category "city" in the DGURBA dataset.
```{r,message = FALSE, results = 'hide'}
gc()
# read city shapefiles ---------------------------------------------------------
# Readme file in INCA's data repository states DEGURBA should be used for 2026
# reporting
cityb<-terra::vect(file.path(dfp,"version_2026_7",
                             "inca_input_local_climate_regulation","Urban_Areas",
                             "2021","DEGURBA",
                             "DGURBA_RG_01M_2021_4258.shp"))
# MM: I checked the data and it has five categories. 1,2 and 3 relate to  degree 
# of urbanisation (1 is cities, 2 is towns and semi-dense areas, 3 is rural). 
# Additionally there is category 9 which exists only in Cyprus, so irrelevant 
# for us. And then there is category 0 which is used in lakes and such. Also not 
# relevant for us. So we subset the data to Norway only.
cityb<-cityb[cityb$CNTR_CODE == "NO",]
# And then we subset to feature category 1 urban only.
cityb<-cityb[cityb$DGURBA == 1,]
# For further operations we reproject the map to EPSG 25833 and we add a 150m
# buffer to the LAU polygons, as suggested by the handbook
cityb <- terra::project(cityb,"EPSG:25833")
cityb_buf <- terra::buffer(cityb,150)
# print the city names
unique(cityb$LAU_NAME)
```
So we get `r length(unique(cityb$LAU_NAME))` LAU in Norway that qualify as city. If we count Sandnes to Stavanger, and Bærum, Rælingen and Lørenskog to Oslo, we have four cities in total, namely Bergen, Oslo, Stavanger, and Trondheim. Let's have a look at the land cover map and street infrastructure for each of the LAU to be considered.
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# read Grunnkartet -------------------------------------------------------------
lcm<-terra::rast(file.path(dfp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))
# project to what we use in this project
lcm <- terra::project(lcm,"EPSG:25833")
terra::plot(lcm, main="Norwegian Land Cover Map (Grunnkart)")

# Make a plot of all cities, possibly simplify Grunnkartet's categorisation

```




Now next we need to identify which of these cities have experienced maximum daily temperatures of at least 25 degrees celsius in the reference year of 2024. For this purpose we first download and import Norwegian meteorological data for 2024 covering the four cities.

MM to revisit the selection of weather stations based on actual LAU shapes.

```{r,message = FALSE, results = 'hide'}
gc()
# read in meteorogogical data for stations inside the above city LAUs for 2024 -
tempd <- read.csv(file.path(here::here(),"ancillary_data",
                            "NO_2024_citylau_dailymaxtemp.csv"),sep=";")
colnames(tempd)<-c("station_name","station_code","date","maxtemp")
tempd$maxtemp <- sub(",",".",tempd$maxtemp )
tempd$maxtemp <- as.numeric(tempd$maxtemp)
# save all cities with daily temperatures above 25 degrees
stations<-tempd %>% filter(maxtemp >= 25) %>% select(station_name) %>% unique()
# check for months in each city
heatmonths<-list()
for (i in 1:nrow(stations)){
  heatmonths[i]<- tempd %>% filter(maxtemp >= 25) %>% filter(station_name==stations[i,1]) %>% select(date) %>% unique()
}
names(heatmonths)<-stations[,1]
for(i in 1:nrow(stations)){
  print(stations[i,1])
  print(unique(substr(heatmonths[[i]],4,5)))
}
```
So this shows that June and July are the months where we have observed daily max temps exceeding 25 degrees in all weather stations within the cities in question in 2024. We therefore pick June and July for the LST data, which we read in next.
```{r}
# read LST raster file ---------------------------------------------------------
# Here we have data available for June, July, August, September, but we only
# pick June and July, as described above.
lstr_6<-terra::rast(file.path(dfp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Land_Surface_Temperature","2024",
                           "06_LST_2024_eu_masked.tif"))
lstr_7<-terra::rast(file.path(dfp,"version_2026_7",
                           "inca_input_local_climate_regulation",
                           "Land_Surface_Temperature","2024",
                           "07_LST_2024_eu_masked.tif"))
# rough crop ext
norcrop <- ext(
  3.5e6, 5.5e6, # xmin, xmax
  3.5e6, 5.5e6 # ymin, ymax
)
lstr_6_no <- crop(lstr_6,norcrop)
lstr_7_no <- crop(lstr_7,norcrop)
# reproject
lstr_6_no <- terra::project(lstr_6_no,"EPSG:25833")
lstr_7_no <- terra::project(lstr_7_no,"EPSG:25833")
# mask to city LAUs
lstr_6_no <- mask(crop(lstr_6_no,cityb_buf),cityb_buf)
lstr_7_no <- mask(crop(lstr_7_no,cityb_buf),cityb_buf)




# Now we reproject and then crop the rasters.
lstr_6<-terra::project(lstr_6,"EPSG:25833")
lstr_7<-terra::project(lstr_7,"EPSG:25833")
















## MM find out what the unit of this is. And need to find out which month of the 
# data we should use. There is something about a scaling factor that should be
# applied. See here: https://www.usgs.gov/landsat-missions/landsat-collection-2-surface-temperature
# MM: what is the day these numbers come from? Or is it something like a monthly
# average?
# need to dig a bit in INCA's GitLab repo and in Lavinia's notes. Possibly also
# NINA's report. Bart at Nina used October. Let's check what Lavinia used.

# Handbook provides guidance on which cities to include. So that is useful.
# if in doubt, just use copernicus data for that. 

# and we should also use a 150m buffer zone to the city LAUs

# see handbook p. 169 - check weather station data for daily max temp. And if
# that is not available, check the copernicus dataset linked in the handbook. 
# this is to inform the selection of months considered in the analysis.

# read TCD raster file ---------------------------------------------------------
lcm<-terra::rast(file.path(dfp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))
terra::plot(lcm, main="Norwegian Land Cover Map (Grunnkart)")

# read EVAP raster file --------------------------------------------------------
lcm<-terra::rast(file.path(dfp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))
terra::plot(lcm, main="Norwegian Land Cover Map (Grunnkart)")

# MMQ then we crop all rasters to the city shapefiles, right? 
```

For illustrative purpose we plot the city shapes here, with the Norwegian land cover map (Grunnkartet) as background. This is to see how much greenspace is present in the cities as defined here.



So in Norway the only LAUs that fall within Eurostat's "city" category are Oslo (with Bærum, Rælingen, Lørenskog), Trondheim, Bergen, and Stavanger (with Sandnes).

## 3.3 Run regressions

We loop across cities here. In each city we run a regression where grid cells are units of observation. Predicting LST with TCD and EVAP. We store the resulting intercepts for each city.

# 4. Estimate a Model of Air Temperature

## 4.1 Overview¨

In this second step we use weather station data to estimate a linear model of maximum air temperature predicted by local surface temperature and latitude. Here is the unit of observation is weather stations. We save the estimated coefficients.

## 4.2 Import data

Read in a single dataframe containing weather station data.

## 4.3 Run regression

A single regression to estimate a linear model. Predicting maximum air temperature with local surface temperature and latitude. The model we estimate is: 
$$ T^{air} = \beta_{2} LST + \gamma_{2}LAT $$

# 5. Simulate Air Temperature for the City Grids

## 5.1 Overview

Here we now apply the estimated coefficients from the previous two steps to generate simulated gridded data. We do this for each city. Given that we have observed LST data at grid level, we compute for each grid cell $i$ the expected air temperature 
$$ T^{air}_{j,i} = \widehat{\beta_{2}} LST_{j,i} + \widehat{\gamma_{2}} latitude_{j} $$. 

Then we compute for each grid cell $i$ the expected air temperature in a no-vegetation scenario, i.e. where $LST_{j,i} = \widehat{\alpha_{1,j}}$. So $$ T^{air,0}_{ji} = \widehat{\beta_{2}} \widehat{\alpha_{1,j}} + \widehat{\gamma_{2}} latitude_{j} $$. 
The cooling effect is then retrieved for each grid cell $i$ in city $j$ as 
$$ cooling_{j,i} = T^{air}_{j,i} - T^{air,0}_{j,i} $$.

## 5.2 Create cooling rasters

Make a loop where we compute the cooling rasters for each city.

# 6. Summarise results

Here we now summarise the data across grid cells (and time).
